\section{Introduction}

This bachelor thesis deals with the formalisation, in the Isabelle/HOL theorem prover, of parts of
the book \emph{Types and Programming Languages}, hereafter abbreviated \emph{TAPL}, by
Benjamen~C.~Pierce. This work concentrate on four of the languages, ranging from simple arithmetic
expressions to the fully fledged lambda-calculus, present in the first two sections, namely "Untyped
Systems" and "Simple Types".

The main motivation to have chosen this subject is the intersection of personal interest and of
opportunities provided by our intership at the chair for logic and verification at TU München.
Having gradually developped an interest for programming languages in the last years, we were eager
to learn more about the foundations of type theory. Pierce's book stood out as a reference
recommended for a deep introduction to the main elements of this field. Also, as part of our
intership, we worked on the implementation of the (Co)datatype module in the Isabelle/HOL theorem
proover. Having experienced the implementator role, we also wanted to learn about the user role and
the process of formalization. The choise of this subject for this thesis was, thus, an opportunity
to fulfill both goals.

Before to dig into the realm of formalizations, we first introduce the required background (section
\ref{sec:background}) in lambda-calculus, type systems and Isabelle/HOL. The lambda-calculus is a
core calculus that captures the essentials features of programming languages. That is, there exists
a way to encode high level features such as recursion, datatypes, records, etc. Such calculus can
come, as do programming languages, in two variant: typed and untyped. A type system is a syntactic
method to prove the absence of certain behaviors such as the misusage of objects of a certain nature
(e.g. natural numbers, booleans, string of characters, etc.) To formalize those we used
Isabelle/HOL, an interactive theorem prover based on higher order logic. It resembles a programming
language in that one can define datatypes and functions. The difference is that it allows to
postulate properties of the formerly defined elements and to provide machine checked poofs that
those properties are theorems.

The formalizations we perform all have a direct corespondance with chapters from \emph{TAPL}
(section \ref{sec:structure-of-formalization}). We provide one Isabelle/HOL theory file per chapter
and usually introduce them in the same order. The only exception is the nameless representation of
terms that we introduce earlier because our formalization depent on it while, in the book, it is an
independant subject.

The untyped arithmetic expressions language (section \ref{sec:untyped-arith-expr}) serves as a
warm-up to experiment with the general structure of formalizations. It consists of boolean
expressions and natural numbers. This simplicity allows to concentrate on the translation to
Isabelle/HOL of the definitions found in the book and to accustoms ourself with the notation. Most
of our definitions and theorems closely follows the ones from the book. The main exceptions being
that we explicitly expose some hypothesis that are implicit in the book and our different definition
of the mutli-step evaluation relation.

The formalization (section \ref{sec:nameless-rep-of-terms}) of the nameless representation of terms,
also known as "de Bruijn indices", was not initially planned but arose from the need to use a
concreate representation for variables in the lambda-calculus. Our formalization closely follows the
book.

While the previous formalizations are either a warm-up or a representation necessity, the untyped
lambda-calculus (section \ref{sec:untyped-lambda-calculus}) is the first core calculus we
formalized. Here, we differ from the book in a non-negligible way. In TAPL, the evaluation
relation assumes that names clashes in variables are automatically solved by renaming them and,
thus, ignore this possibility from there on. Such an assumption is not accepted by computer-verified
proof. We choosed to use "de Bruijn indices" as representation for variable to encode this
assumption. Also, since the chapter in the book is more focused on explaining the lambda calculus,
it does not contain meaningful theorems. Nevertheless, we reprove, or disprove, the theorems
introduced with the arithmetic expressions language.

The typed arithmetic expressions language (section \ref{sec:typed-arith-expr}) is again a warm-up;
this time, to experiment with the formalization of a type system. Our formalization closely follows
the book.

The simply typed lambda-calculus (section \ref{sec:simply-typed-lambda-calculus}) is the second core
calculus we formalize. Here, we differ significantly from the book. Mainly because of our use of "de
Bruijn indices" but also because of our reprensentation of the typing context, we need to adapt some
lemmas and replace others. This is certainly the most challenging part of the formalization, since
we can not follow the described proofs anymore and must find the right assumptions for our lemmas.
In spite of these differences, we belive that our formalization still respect the spirit of the book
since only the helper lemmas chang and the theorems remain the same.

All sections combined, the formalization consists of N lines of definitions, theorems and exercies
proposed in the book. It is publicly available
\footnote{https://github.com/authchir/log792-type-systems-formalization} and can be executed with
Isabelle/HOL 2014\footnote{http://isabelle.in.tum.de/website-Isabelle2014/}. In this report, we
focus on the definitions and how the theorems are expressed. When relevant, we present both the
definitions from the book and our translation, highlighting and motivating the differences. Some
proofs are presented but not explained. For a deeper insight on the proofs, the best methodology is
to study the theory files in Isabelle.

% $ cat *.thy | wc
%    1728   10170   68275

% sed -nr '1h;1!H;${;g;s/\(\*.*\*\)//g;p;}' test.thy
